Lab 9: Multinomial and Ordered Probit Models: Answers

### 2023-03-21 Lab 9  Multinom_orderedprob/Lab9_Multinom_OrderedProbit.Rmd
Add files via upload

Load necessary packages:
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, messsage = FALSE)
rm(list=ls())

#Load in packages
pacman::p_load(tidyverse,
               haven,#for importing data
               labelled,
               MASS,
               arm)

@@

For ordered outcomes, we need to use the MASS package (rather than base R). Otherwise, our syntax looks much the same.

@@

#Now, our model:
m1 <- polr(apply ~ pared + public + gpa, 
           data = grad, method = "probit", Hess = TRUE)
summary(m1)

#What is the effect of increasing each predictor by one unit?
#How do the odds of moving from a lower to adjacent higher category change?
coefs <- m1$coefficients[1:3]
#odds ratios
exp(coefs)

@@

We can also check this first: 
```{r}
tau1 <- m1$zeta[1] #extract latent thresholds (calculated w polr) from our model!
tau2 <- m1$zeta[2]
p1 <- invlogit(tau1) # 1
p2 <- invlogit(tau2) - invlogit(tau1) # 2
p3 <- 1-invlogit(tau2)

@@

```{r}
# if we wanted to plot the range of gpa for all four combinations of pared and
#public, we could do the following:
#grad_2 <- expand.grid(pared = c(0:1), 
#                      public = c(0,1), 
#                      gpa = seq(from = 1.9, to = 4, length.out = 1000))
#We would then create an indicator variable for the four combinations, and then
#use facet_wrap in ggplot to create 4 different plots
#however, if we just want one easily interpretable plot, we can do the following:
table(grad$pared); table(grad$public)
grad_2 <- expand.grid(pared = 0, #modal value
                      public = 0, #modal value
                      gpa = seq(from = 1.9, to = 4, length.out = 1000))
colnames(grad_2)<- c("pared", "public", "gpa")
probs <- predict(m1, grad_2, type = "probs") #type = is different than for binomial model
head(rowSums(probs)) #just to double check
cumul_probs <- t(apply(probs, 1, cumsum)) #get the cumulative sum of the rows
#transpose into columns for plotting and merging with values of gpa
head(cumul_probs)
plot_data <- data.frame(grad_2$gpa, cumul_probs)
ggplot(data=plot_data, aes(x=grad_2.gpa))+
    geom_smooth(data=plot_data, aes(x=grad_2.gpa, y=unlikely), se=FALSE)+
    geom_smooth(data=plot_data, aes(x=grad_2.gpa, y=somewhat.likely), se=FALSE)+
    geom_smooth(data=plot_data, aes(x=grad_2.gpa, y=very.likely), se=FALSE)+
    geom_ribbon(data=plot_data, aes(ymin=0, ymax=unlikely, fill="Unlikely"))+
    geom_ribbon(data=plot_data, aes(ymin=unlikely, ymax=somewhat.likely,
                                    fill="Somewhat Likely"))+
    geom_ribbon(data=plot_data, aes(ymin=somewhat.likely, ymax=very.likely,
                                    fill="Very Likely"))+
    labs(x="Values of GPA", y="Cumulative Predicted Probability")+
    scale_fill_discrete(name="Applicant Response")+
    theme_bw()
```

@@

coef_samp <- mvrnorm(1e3,coef(m1)
                     ,vcov(m1)[1:3,1:3]) #keep relevant vals of vcov matrix

## Get linear predictors
hyp_X <- model.matrix(~pared+public+gpa
                      ,data = grad_3)[,-1] #no intercept 
eta <- hyp_X %*% t(coef_samp) 
summary(t(eta)) 
## Get pred probs samples 
taus <- m1$zeta #extract thresholds 
pred_prob_diffs <- apply(eta,c(1,2)
                    ,function(x, J, tau){
                      prob_vec <- array(NA,J) #J categories - array() is generalizable
                      prob_vec[1] <- plogis(tau[1]-x) #prob of being in 1st category
                      for(i in 2:(J-1)) #move to next category
                        prob_vec[i] <- plogis(tau[i]-x) - plogis(tau[i-1]-x)
                        #probability of being in middle category
                      prob_vec[J] <- 1 - plogis(tau[J-1]-x)
                        #probability of being in top category (sums to 1)
                      return(prob_vec[J]-prob_vec[1]) 
                      # Diff. in prob from top to bottom cat.
                    }
                    ,J = 3 #3 possible outcomes
                    ,tau = taus)

@@

pacman::p_load(nnet,#for multinom()
               mlogit,#for conditional mnl
               cowplot)#we will use it to put plots together (an extension to ggplot2)

@@

both <- plot_grid(mplot, fplot, labels = c('Male', 'Female'))
both

@@

both2 <- plot_grid(cap_plot, nocap_plot, labels = c('Cap: Yes', 'Cap: No'))
both2

@@

#Create the model
warm_model <- polr(____, Hess = TRUE)
summary(warm_model)

#What is the effect of increasing each predictor by one unit?
#How do the odds of moving from a lower to adjacent higher category change?
coefs <- warm_model$_____
#odds ratios
_____(coefs)

@@

probs <- predict(____, type = "probs") #type = is different than for binomial model

cumul_probs <- t(apply(probs, 1, cumsum)) #get the cumulative sum of the rows
#transpose into columns for plotting and merging with values of gpa

>>>>

Load necessary packages:
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, messsage = FALSE,warning = FALSE)
rm(list=ls())

#Load in packages
pacman::p_load(tidyverse,
               haven,#for importing data
               labelled,
               ordinal,
               arm)

@@

For ordered outcomes, we need to use the ordinal package (rather than base R). Otherwise, our syntax looks much the same.

@@

#Now, our model:
m1 <- clm(apply ~ pared + public + gpa, 
           data = grad, link = "probit")
summary(m1)

#What is the effect of increasing each predictor by one unit?
#How do the odds of moving from a lower to adjacent higher category change?
coefs <- m1$coefficients[3:5]
#odds ratios
exp(coefs)

@@

We can also check this first: 
```{r}
tau1 <- m1$alpha[1] #extract latent thresholds (calculated w polr) from our model!
tau2 <- m1$alpha[2]
p1 <- invlogit(tau1) # 1
p2 <- invlogit(tau2) - invlogit(tau1) # 2
p3 <- 1-invlogit(tau2)

@@

```{r}
# if we wanted to plot the range of gpa for all four combinations of pared and
#public, we could do the following:
#grad_2 <- expand.grid(pared = c(0:1), 
#                      public = c(0,1), 
#                      gpa = seq(from = 1.9, to = 4, length.out = 1000))
#We would then create an indicator variable for the four combinations, and then
#use facet_wrap in ggplot to create 4 different plots
#however, if we just want one easily interpretable plot, we can do the following:
table(grad$pared); table(grad$public)
grad_2 <- expand.grid(pared = 0, #modal value
                      public = 0, #modal value
                      gpa = seq(from = 1.9, to = 4, length.out = 1000))
colnames(grad_2)<- c("pared", "public", "gpa")
probs <- predict(m1, grad_2, type = "prob") #type = is different than for binomial model

#?predict.clm

head(rowSums(probs$fit)) #just to double check
cumul_probs <- t(apply(probs$fit, 1, cumsum)) #get the cumulative sum of the rows
#transpose into columns for plotting and merging with values of gpa
head(cumul_probs)
plot_data <- data.frame(grad_2$gpa, cumul_probs)
ggplot(data=plot_data, aes(x=grad_2.gpa))+
    geom_smooth(data=plot_data, aes(x=grad_2.gpa, y=unlikely), se=FALSE)+
    geom_smooth(data=plot_data, aes(x=grad_2.gpa, y=somewhat.likely), se=FALSE)+
    geom_smooth(data=plot_data, aes(x=grad_2.gpa, y=very.likely), se=FALSE)+
    geom_ribbon(data=plot_data, aes(ymin=0, ymax=unlikely, fill="Unlikely"))+
    geom_ribbon(data=plot_data, aes(ymin=unlikely, ymax=somewhat.likely,
                                    fill="Somewhat Likely"))+
    geom_ribbon(data=plot_data, aes(ymin=somewhat.likely, ymax=very.likely,
                                    fill="Very Likely"))+
    labs(x="Values of GPA", y="Cumulative Predicted Probability")+
    scale_fill_discrete(name="Applicant Response")+
    theme_bw()
```

@@

coef_samp <- mvrnorm(1e3,coef(m1)[3:5]
                     ,vcov(m1)[3:5,3:5]) #keep relevant columns

## Get linear predictors
hyp_X <- model.matrix(~pared+public+gpa
                      ,data = grad_3)[,-1] #no intercept 
eta <- hyp_X %*% t(coef_samp) 
summary(t(eta)) 
## Get pred probs samples 
taus <- m1$alpha #extract thresholds 
pred_prob_diffs <- apply(eta,c(1,2)
                    ,function(x, J, tau){
                      prob_vec <- array(NA,J) #J categories - array() is generalizable
                      prob_vec[1] <- plogis(tau[1]-x) #prob of being in 1st category
                      for(i in 2:(J-1)) #move to next category
                        prob_vec[i] <- plogis(tau[i]-x) - plogis(tau[i-1]-x)
                        #probability of being in middle category
                      prob_vec[J] <- 1 - plogis(tau[J-1]-x)
                        #probability of being in top category (sums to 1)
                      return(prob_vec[J]-prob_vec[1]) 
                      # Diff. in prob from top to bottom cat.
                    }
                    ,J = 3 #3 possible outcomes
                    ,tau = taus)

@@

pacman::p_load(nnet,#for multinom()
               mlogit,#for conditional mnl
               patchwork)#we will use it to put plots together (an extension to ggplot2)

@@

#Combining plot using patchwork (could also use cowplot())
both<-mplot+fplot

#Adding annotations
both<-both + plot_annotation(tag_levels = 'a')

both<-both + plot_annotation(tag_levels = list(c("Male","Female")))& 
  theme(plot.tag.position = c(0, 1),
        plot.tag = element_text(hjust = -.5, vjust = -0.20))

both

@@

#combining
both2<-cap_plot+nocap_plot

#Adding annotations
both2<-both2 + plot_annotation(tag_levels = 'a')

both2<-both2 + plot_annotation(tag_levels = list(c("Cap: Yes","Cap: No")))& 
  theme(plot.tag.position = c(0, 1),
        plot.tag = element_text(hjust = -.5, vjust = -0.20))

both2

@@

#Create the model
warm_model <- clm(____,)
summary(warm_model)

#What is the effect of increasing each predictor by one unit?
#How do the odds of moving from a lower to adjacent higher category change?
coefs <- warm_model$coefficients[5:8]
#odds ratios
_____(coefs)

@@

probs <- predict(____, type = "prob") #type = is different than for binomial model

cumul_probs <- t(apply(probs$fit, 1, cumsum)) #get the cumulative sum of the rows
#transpose into columns for plotting and merging with values of gpa

### 2023-03-21 Lab 9  Multinom_orderedprob/Lab9_Multinom_OrderedProbit_Answers.Rmd
Add files via upload

#Load in packages
pacman::p_load(tidyverse,
               haven,#for importing data
               labelled,
               MASS,
               arm)

@@

We can also check this first: 
```{r}
tau1 <- m1$zeta[1] #extract latent thresholds (calculated w polr) from our model!
tau2 <- m1$zeta[2]
p1 <- invlogit(tau1) # 1
p2 <- invlogit(tau2) - invlogit(tau1) # 2
p3 <- 1-invlogit(tau2)

@@

#Create the model
warm_model <- polr(warm~male+white+age+ed+prst, 
           data = hw, method = "probit", Hess = TRUE)
summary(warm_model)

#What is the effect of increasing each predictor by one unit?
#How do the odds of moving from a lower to adjacent higher category change?
coefs <- warm_model$coefficients[1:5]
#odds ratios
exp(coefs)

@@

probs <- predict(warm_model, hw_2, type = "probs") #type = is different than for binomial model
head(rowSums(probs)) #just to double check
cumul_probs <- t(apply(probs, 1, cumsum)) #get the cumulative sum of the rows
#transpose into columns for plotting and merging with values of gpa
head(cumul_probs)
plot_data <- data.frame(hw$prst, cumul_probs)

>>>>

#Load in packages
pacman::p_load(tidyverse,
               haven,#for importing data
               labelled,
               ordinal,
               arm)

@@

We can also check this first: 
```{r}
tau1 <- m1$alpha[1] #extract latent thresholds (calculated w clm) from our model!
tau2 <- m1$alpha[2]
p1 <- invlogit(tau1) # 1
p2 <- invlogit(tau2) - invlogit(tau1) # 2
p3 <- 1-invlogit(tau2)

@@

#Create the model
warm_model <- clm(warm~male+white+age+ed+prst, 
           data = hw, link = "probit")
summary(warm_model)

#What is the effect of increasing each predictor by one unit?
#How do the odds of moving from a lower to adjacent higher category change?
coefs <- warm_model$coefficients[5:8]
#odds ratios
exp(coefs)

@@

probs <- predict(warm_model, hw_2, type = "prob") #type = is different than for binomial model
head(rowSums(probs$fit)) #just to double check
cumul_probs <- t(apply(probs$fit, 1, cumsum)) #get the cumulative sum of the rows
#transpose into columns for plotting and merging with values of gpa
head(cumul_probs)
plot_data <- data.frame(hw$prst, cumul_probs)
